---
title: "TFCB Capstone Question 1"
author: "Amin Addetia"
date: "12/15/2020"
output: html_document
---

# Impact of age and glaucoma status on candidate features used to develop a machine learning model for diagnosing glaucoma

The dataset analyzed here examines various features in individuals diagnosed with glaucoma compared to healthy individuals and then builds uses these data to build a machine learning model for diagnosing glaucoma. The model asssumes that each the of the factors are equally predictive regardless of the individual's age and uses the individual's age in the machine learning diagnosis decision tree. Here, I examine that underlying assumption by asking the following question: Are certain measured factors better predictors of glaucoma status for particular age groups or are all factors equally good predictors of glaucoma status regardless of an individual's age?


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Loading the relevant libraries for the analysis.

library(tidyverse)
library(ggpubr)

```

## Reading in dataset and generating summary statistics for groups stratified by age and glaucoma status

This section reads in the [ds_whole.csv](data/ds_whole.csv) file and outputs the summary file [age_group_glaucoma_summary_statistics.csv](age_group_glaucoma_summary_statistics.csv).

```{r cars}

#Reading in dataset

dataset <- read.csv('data/ds_whole.csv', header = T)

#Reassigning variables in glaucoma column to make more human readable

dataset %>%
  mutate(glaucoma=case_when(glaucoma == 0 ~ 'No',
                   glaucoma == 1 ~ 'Yes'), .keep = 'unused') -> dataset

#Adding a new column with age group for each observation, which is based on the age column

dataset %>% 
  mutate(age_group=case_when(age <= 20 ~ '20 yo and under',
                             age >= 21 & age <= 35 ~ '21-35 yo',
                             age >= 36 & age <= 50 ~ '36-50 yo',
                             age >= 51 & age <= 65 ~ '51-65 yo',
                             age >= 66 & age <= 80 ~ '66-80 yo',
                             age > 80 ~ '80 yo and over')) -> dataset

#Summarize the data based on glaucoma status and age group

dataset %>%
  group_by(age_group, glaucoma) %>%
  summarise_all(funs(mean)) -> summary

#Dropping the RL column

summary <- subset(summary, select = -c(RL))

#Renaming the column headers to make them more human-readable

summary %>%
  rename( 'Age Group'= age_group,
          Glaucoma = glaucoma,
          'Mean Age' = age,
          'Mean ocular pressure' = ocular_pressure,
          'Mean MD' = MD,
          'Mean PSD' = PSD,
          'Mean GHT' = GHT,
          'Mean cornea thickness' = cornea_thickness,
          'Mean RNFL thickness' = RNFL4.mean) -> summary

#Displaying the summary statistics

print(summary)

#Creating a csv file with the summary statistics

write.csv(summary, 'age_group_glaucoma_summary_statistics.csv', row.names = F)

```

## Plotting data stratified by age group and glaucoma.

This section takes the summary table generated in the chunk above and outputs plots for each of the candidate features in the [age_group_glaucoma_plots.png](age_group_glaucoma_plots.png) file.

```{r pressure, echo=FALSE}

#Creating individual boxplots for each of the variables, grouped by age group and glaucoma status.

MD_plot <- ggplot(dataset, aes(x=age_group, y=MD, fill=factor(glaucoma))) + geom_boxplot() + theme_classic() + xlab('Age Group') + labs(fill='Glaucoma') + theme(axis.text.x = element_text(angle = 90))
 
OP_plot <- ggplot(dataset, aes(x=age_group, y=ocular_pressure, fill=factor(glaucoma))) + geom_boxplot() + theme_classic() + xlab('Age Group') + labs(fill='Glaucoma') + ylab('Ocular Pressure') + theme(axis.text.x = element_text(angle = 90))
 
PSD_plot <- ggplot(dataset, aes(x=age_group, y=PSD, fill=factor(glaucoma))) + geom_boxplot() + theme_classic() + xlab('Age Group') + labs(fill='Glaucoma') + theme(axis.text.x = element_text(angle = 90))
 
GHT_plot <- ggplot(dataset, aes(x=age_group, y=GHT, fill=factor(glaucoma))) + geom_boxplot() + theme_classic() + xlab('Age Group') + labs(fill='Glaucoma') + theme(axis.text.x = element_text(angle = 90))

CT_plot <- ggplot(dataset, aes(x=age_group, y=cornea_thickness, fill=factor(glaucoma))) + geom_boxplot() + theme_classic() + xlab('Age Group') + labs(fill='Glaucoma') + ylab('Cornea Thickness') + theme(axis.text.x = element_text(angle = 90))

RNFL4_plot <- ggplot(dataset, aes(x=age_group, y=RNFL4.mean, fill=factor(glaucoma))) + geom_boxplot() + theme_classic() + xlab('Age Group') + labs(fill='Glaucoma') + ylab('RNFL Mean Thickness') + theme(axis.text.x = element_text(angle = 90))

#Arranging all the individual plots together in a single plot

plots <- ggarrange(MD_plot, OP_plot, PSD_plot, GHT_plot, CT_plot, RNFL4_plot, common.legend = T, legend = 'bottom') + theme(axis.text.x = element_text(angle = 90))

#Displaying the final plots

plot(plots)

#Saving the final plots as a png file

ggsave('age_group_glaucoma_plots.png', plot = plots)


```

To answer my research question, I visualized the data as boxplots stratified by age group and glaucoma status. I chose to use box plots for the plots as they display both the mean and range for the plotted dataset. As a result, they allow for an easy visual comparison of distributions between the glaucoma-positive and glaucoma-negative groups in each age group. I kept the coloring of glaucoma status identical in each plot for easy visual comparison across the plots.

The summary table and plots indicate that certain factors might be predictors of glaucoma status for particular age groups. For example, MD is likely a better predictor of glaucoma status for the 80 years old and over group than for other age groups. Further, ocular pressure is likely a better predictor of glaucoma status for younger individual than for older individuals. 

# Glaucoma diagnosis predicted by RNFL mean thickness

In the machine learning decision tree, the first factor examined for used for determining an individual's glaucoma status is the RNFL mean thickness. The authors used a RNFL mean thickness of 89.33334 as their initial criteria in their glaucoma decision tree. Here, I examine how many individuals would be correctly diagnosed by RNFL mean thickness alone and ask: Is the chosen RNFL mean thickness value the best value for distinguishing between the glaucoma-positive and glaucoma-negative individuals or does a better value exist?


## Calculating the number of individuals correctly and incorrectly diagnosis on RNFL mean thickness value alone.

This section filters the dataset to determine how many individuals can be correctly diagnosed with RNFL mean thickness alone and calculates the false positive and false negative rates.

```{r}

#Counting number of individuals correctly predicted to not have glaucoma and calculating the false negative rate

dataset %>%
  filter(RNFL4.mean > 89.33334) %>%
  group_by(glaucoma) %>%
  count() %>%
  ungroup() %>%
  mutate(Percent=n/sum(n)*100) -> RNFL_no_glaucoma_89

#Counting number of individuals correctly predicted to  have glaucoma and calculating the false positive rate

dataset %>%
  filter(RNFL4.mean <= 89.33334) %>%
  group_by(glaucoma) %>%
  count() %>%
  ungroup() %>%
  mutate(Percent=n/sum(n)*100) -> RNFL_glaucoma_89

#Printing the results

print(paste0(RNFL_glaucoma_89[2,2], ' individuals were correctly predicted to have glaucoma and ',  RNFL_no_glaucoma_89[1,2], ' individual were correctly to predicted to not have glaucoma using a RNFL4 mean thickness of 89.3334. The false positive rate was ', RNFL_glaucoma_89[1,3], ' and the false negative rate was ', RNFL_no_glaucoma_89[2,3], '.'))
```


```{r}

#Repeating analysis with RNFL mean thickness of 87

dataset %>%
  filter(RNFL4.mean > 87) %>%
  group_by(glaucoma) %>%
  count() %>%
  ungroup() %>%
  mutate(Percent=n/sum(n)*100) -> RNFL_no_glaucoma_87

dataset %>%
  filter(RNFL4.mean <= 87) %>%
  group_by(glaucoma) %>%
  count() %>%
  ungroup() %>%
  mutate(Percent=n/sum(n)*100) -> RNFL_glaucoma_87

print(paste0(RNFL_glaucoma_87[2,2], ' individuals were correctly predicted to have glaucoma and ',  RNFL_no_glaucoma_87[1,2], ' individual were correctly to predicted to not have glaucoma using a RNFL4 mean thickness of 87. The false positive rate was ', RNFL_glaucoma_87[1,3], ' and the false negative rate was ', RNFL_no_glaucoma_87[2,3], '.'))
```


```{r}

#Repeating analysis with RNFL mean thickness of 91

dataset %>%
  filter(RNFL4.mean > 91) %>%
  group_by(glaucoma) %>%
  count() %>%
  ungroup() %>%
  mutate(Percent=n/sum(n)*100) -> RNFL_no_glaucoma_91

dataset %>%
  filter(RNFL4.mean <= 91) %>%
  group_by(glaucoma) %>%
  count() %>%
  ungroup() %>%
  mutate(Percent=n/sum(n)*100) -> RNFL_glaucoma_91

print(paste0(RNFL_glaucoma_91[2,2], ' individuals were correctly predicted to have glaucoma and ',  RNFL_no_glaucoma_91[1,2], ' individual were correctly to predicted to not have glaucoma using a RNFL4 mean thickness of 91. The false positive rate was ', RNFL_glaucoma_91[1,3], ' and the false negative rate was ', RNFL_no_glaucoma_91[2,3], '.'))
```


## Plotting the distribution of RNFL mean thickness for glaucoma-positive and glaucoma-negative individuals

This section plots the RNFL mean thickness stratified by glaucoma status to examine why choosing a RNFL mean thickness greater or less than 89.33334 increased the false positive or false negative rates, respectively. 

```{r}

#Creating density plot

RNFL_density_plot <- ggplot(dataset, aes(x = RNFL4.mean, color = glaucoma, fill = glaucoma)) + geom_density(alpha = 0.2) + theme_classic() + geom_vline(xintercept = 89.33334, linetype = 'dashed') + xlab('RNFL mean thickness')

#Displaying plot

plot(RNFL_density_plot)

#Saving the plot

ggsave('RNFL_mean_thickness_density_plot.png')

```
I chose to represent the distribution RNFL mean thickness as a density plot as it allowed for an easy visual examination of the distributions of the glaucoma-positive and glaucoma-negative groups. I kept the color identical to the previous plots for easy visual interpretation. 

The analysis here shows that increasing the RNFL mean thickness for  would increase the false positive rate while decreasing the RNFL mean thickness would increase the false negative rate. This is further demonstrated in the density plot which shows shifting the chosen RNFL mean thickness value impacts the number of false positive or false negative cases since the distribution overlap. Together, this demonstrates that RNFL mean thickness alone is not a perfect predictor of glaucoma status.
